---
title:  "HELPmiss (Data Verbs) Activity"
author: "Dave Friedman"
output: html_notebook
---

## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData) # includes the `HELPmiss` data set 

# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")
```

## Helpful links: 

- Look though the DC Textbook for `tidyverse` functions. 
- Check out some of these RStudio cheat sheets:
  - <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf> 
  - <https://www.rstudio.com/resources/cheatsheets/>
  - <https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf>

## Task 1: Data Description

*Write several sentences (or a bullet list) describing the HELP Study and the resulting `HELPmiss` data.  Your description should investigate basic data provenance (e.g., Who, What, When, Where, Why, How), explain the setting for the data, specify what each case represents in the data, and remark on inclusion/exclusion criteria.* 

- investigate by searching R help documentation
- there's a research paper cited where additional detail is provided

"The HELP study was a clinical trial for adult inpatients recruited
from a detoxification unit.  Patients with no primary care
physician were randomized to receive a multidisciplinary
assessment and a brief motivational intervention or usual care,
with the goal of linking them to primary medical care.

Eligible subjects were adults, who spoke Spanish or English,
reported alcohol, heroin or cocaine as their first or second drug
of choice, resided in proximity to the primary care clinic to
which they would be referred or were homeless. Patients with
established primary care relationships they planned to continue,
significant dementia, specific plans to leave the Boston area that
would prevent research participation, failure to provide contact
information for tracking purposes, or pregnancy were excluded.

Subjects were interviewed at baseline during their detoxification
stay and follow-up interviews were undertaken every 6 months for 2
years.  A variety of continuous, count, discrete, and survival
time predictors and outcomes were collected at each of these five
occasions.

This dataset is a superset of the HELPrct data with 17 subjects
with partially observed data on some of the baseline variables.
This is a subset of the HELPfull data which includes 5 timepoints
and many additional variables."

## Task 2: Basic Summaries 

*Produce one or more R expressions involving `summarize()` and `HELPmiss` to address each of the following prompts.*

1. number of people (cases) in `HELPmiss` study

```{r}
HELPmiss %>%
  summarize(
    count = n()
  )
```

2. total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss`. 

```{r}
HELPmiss %>%
  summarize(
    total_e2b = sum(e2b, na.rm=TRUE)
  )
```

3. mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`

```{r}
HELPmiss %>%
  summarize(
    mean_daysanysub = mean(daysanysub, na.rm=TRUE)
  )
```

## Task 3: Group Summaries  

*Repeat task 2 above, but add code chunks to calculate result group-by-group according to each prompt (i.e. each promt should have three statements for each of the three prompts in task 2).  Be sure to show all R code and write a sentence or two about what you observe in the results. Remember, you can add multiple statistics inside `summary`.* 

- males versus females
- homeless or not
- substance
- break down the homeless versus housed further, by sex
- homeless versus housed broken down by substance

### males versus females
```{r}
HELPmiss %>%
  group_by(sex) %>%
    summarize(
      count           = n(),
      total_e2b       = sum(e2b, na.rm=TRUE),
      mean_daysanysub = mean(daysanysub, na.rm=TRUE)
    )
```

### homeless or not
```{r}
HELPmiss %>%
  group_by(homeless) %>%
    summarize(
      count           = n(),
      total_e2b       = sum(e2b, na.rm=TRUE),
      mean_daysanysub = mean(daysanysub, na.rm=TRUE)
    )
```

### substance
```{r}
HELPmiss %>%
  group_by(substance) %>%
    summarize(
      count           = n(),
      total_e2b       = sum(e2b, na.rm=TRUE),
      mean_daysanysub = mean(daysanysub, na.rm=TRUE)
    )
```

### homeless versus housed broken down by sex
```{r}
HELPmiss %>%
  group_by(homeless, sex) %>%
    summarize(
      count           = n(),
      total_e2b       = sum(e2b, na.rm=TRUE),
      mean_daysanysub = mean(daysanysub, na.rm=TRUE)
    )
```

### homeless versus housed broken down by substance
```{r}
HELPmiss %>%
  group_by(homeless, substance) %>%
    summarize(
      count           = n(),
      total_e2b       = sum(e2b, na.rm=TRUE),
      mean_daysanysub = mean(daysanysub, na.rm=TRUE)
    )
```

## Task 4: Data Visualization & Observations

*Include one or more interesting plots from this data set involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can expand on the relationships that you studied in Task 2, or you can explore a different group of variables in `HELPmiss` that show something interesting. Remember to use the interactive commands in the console, generate the R commands that will reproduce your plot, and then paste the R commands into an R chunk in the RMarkdown file.*  

*Remember, you can use* `esquisser` *or* `mplot` *in your console. But only include the ggplot code in this Rmd document.*
 
Most of the data pertain to black and white males. It appears as though black males are most likely to abuse cocaine and white males are most likely to abuse alcohol over the course of their life. The same trend arguably exists for females.

```{r}
ggplot(data = HELPmiss) +
  geom_bar(
    mapping  = aes(x    = age,
                   fill = substance),
    position = 'stack',
    stat     = 'count',
    show.legend=TRUE
  ) +
  facet_grid(sex ~ racegrp)
```